Chatbots and messaging platforms in the classroom: an analysis from the teacher's perspective

### 2021-07-16 edubots_surveys.Rnw
Latest comments and suggestions by @noorhanabbas @aguillen @olitsi
sorts things out in RQ1 and adds tables

For answering RQ1, teachers were queried about whether they use messaging apps in their classes, specifically, Telegram, WhatsApp, Slack, or any other messaging app, and whether they use messaging apps provided by their academic institution. Overall, the majority of the teachers responded that they do use messaging apps in their classrooms, from which apps provided by the academic institution (n=159, 56\%) and WhatsApp (n=124, 44.0\%) were the most common responses. Only 19 teachers (6.7\%) replied that they do not use any messaging app in their class.

% Insert first Olia table here
A chi-square test of independence was performed to examine the relation between instructors’ discipline, sector, and gender and any potential changes that occurred to the use of messaging apps due to the covid-19 pandemic. The relation between the latter variable and the instructors’ sector was significant, X2 (4, N = 282) = 9.598, p =0.048. The frequencies cross tabulated are given in Table 2. In effect, this finding indicates that how teachers responded to the use of messaging apps during the pandemic and in particular, whether they changed their habits in the use of apps for teaching purposes, was related to their sector (university vs vocational). 
Although gender differences are not significant, female teachers answered they use instant messaging apps more than male teachers (about 10\% more). Also, teachers in vocational Education also use Whatsapp more than university teachers.
%Insert second Olia table here.

With respect to the distribution of the use of messaging apps per age, there are no significant differences for Whatsapp and apps provided by their institutions. % Olia requests support for this.
However, younger teachers use also Telegram with more than 25\%, a percentage that falls to about 10\% for teachers that are 55 or older. The third column compares the use of these apps considering the number of years of experience. One interesting result is that, contrarily to the idea of technology adaptation issues when considering older teachers, about 65\% of teachers with more than 25 years of experience use the platforms provided by their institutions while the percentage goes down to less than 50\% for teachers with 6-15 years of experience.

@@

Some of the questions in the survey were focused on the impact of the COVID19 pandemic between 2020 and 2021, in the teachers' opinions towards the instant messaging apps. The 273 % Fran, please check Olia's comment as having 282 answers.
answers are summarized in Fig. \ref{figure:chatbots_use_postcovid}, showing that about 77\% of teachers already used these tools before the pandemic and kept using them during the pandemic lockdowns that forced students and educators to use remote education schemes. Moreover, approximately 15\% of them switched their messaging app for one that offered a safer interaction with their students. According to the responses, an additional 16\% started using messaging apps during the pandemic for the first time in their classes. Aggregating the answers, currently 91\% of the teachers are using messaging apps as learning tools.

@@

\begin{figure}[h!tbp]
\begin{center}
    \begin{minipage}[t]{0.85\textwidth}
        \centering
<<gender_covid, echo=FALSE, fig.width=19/2.54, fig.height=5.5/2.54>>=
survey.pilot.postCOVID <- within(survey.pilot.postCOVID, PostCOVID.changes <- factor(PostCOVID.changes, levels=names(sort(table(PostCOVID.changes), decreasing=TRUE))))
ggplot(survey.pilot.postCOVID, aes( x = PostCOVID.changes, y = ..count.., group=Gender, fill = Gender ) ) + geom_bar()+ theme(axis.text.x = element_text(angle = 90)) +coord_flip()+ theme_tufte()+ theme(axis.title.y=element_blank())
@
%       \includegraphics[width=\textwidth]{figures/gender_covid.pdf}
    \end{minipage}
    \begin{minipage}[t]{0.85\textwidth}
        \centering
<<sector_covid, echo=FALSE, fig.width=19/2.54, fig.height=5.5/2.54>>=
ggplot(survey.pilot.postCOVID, aes( x = PostCOVID.changes, y = ..count.., group=Sector, fill = Sector ) ) + geom_bar()+ theme(axis.text.x = element_text(angle = 90)) +coord_flip()+ theme_tufte()+ theme(axis.title.y=element_blank())
@
%       \includegraphics[width=\textwidth]{figures/sector_covid.pdf}
    \end{minipage}
    \begin{minipage}[t]{0.85\textwidth}
        \centering
<<experience_covid, echo=FALSE, fig.width=19/2.54, fig.height=5.5/2.54>>=
ggplot(survey.pilot.postCOVID, aes( x = PostCOVID.changes, y = ..count.., group=Experience, fill = Experience ) ) + geom_bar()+ theme(axis.text.x = element_text(angle = 90)) +coord_flip()+ theme_tufte()+ theme(axis.title.y=element_blank())
@       
%       \includegraphics[width=\textwidth]{figures/experience_covid.pdf}
    \end{minipage}
\end{center}
    \caption{Impact of COVID in the use of messaging apps in University and vocational education.}
\label{figure:chatbots_use_postcovid}
\end{figure}

A chi-square test of independence was performed to examine the relation between instructors’ discipline, sector, and gender and any potential changes that occurred to the use of messaging apps due to the covid-19 pandemic. The relation between the latter variable and the instructors’ sector was significant, $X^{2}_{(4, N = 282)} = 9.598$, $p =0.048$. The frequencies cross tabulated are given in Table \ref{tab:sector*covid}. In effect, this finding indicates that how teachers responded to the use of messaging apps during the pandemic and in particular, whether they changed their habits in the use of apps for teaching purposes, was related to their sector (university vs vocational). The relationships between changes in the use of messaging apps, due to the covid-19 pandemic, and the instructors’ discipline, $X^{2}_{(24, N = 282)} = 44.856$, $p =0.006$, as well as the gender, $X^{2}_{(8, N = 282)} = 16.249$, $p= 0.039$ were also significant.

%TODO: This table is not referenced in the text
\begin{table}[h!tbp]
\begin{tabular}{lllll}
\hline
\multicolumn{1}{c}{Messaging App} & \multicolumn{2}{c}{Yes} & \multicolumn{2}{l}{No} \\ \cline{2-5} 
\multicolumn{1}{c}{} & \multicolumn{1}{c}{Frequency} & \multicolumn{1}{c}{\%} & Frequency & \multicolumn{1}{c}{\%} \\ \hline
Telegram & 62 & 22.0 & 220 & 78.0 \\
WhatsApp & 124 & 44.0 & 158 & 56.0 \\
Slack & 15 & 5.3 & 267 & 94.7 \\
Other & 60 & 21.3 & 222 & 78.7 \\
\begin{tabular}[c]{@{}l@{}}Provided by the Academic\\ Institution\end{tabular} & 159 & 56.4 & 123 & 43.6 \\
None & 19 & 6.7 & 263 & 93.3 \\ \hline
\end{tabular}
\vspace{0.8em}
\caption{Use of messaging apps in class}
\label{tab:use_messapps}
\end{table}

\begin{table}[h!tbp]
\begin{tabular}{llccc}
\hline
 & \multicolumn{1}{c}{} & \multicolumn{2}{c}{Sector} & Total \\ \cline{3-5} 
 & \multicolumn{1}{c}{} & \begin{tabular}[c]{@{}c@{}}University \\ (f)\end{tabular} & \begin{tabular}[c]{@{}c@{}}Vocational\\ (f)\end{tabular} &  \\ \cline{2-5} 
\multirow{4}{*}{\begin{tabular}[c]{@{}l@{}}Post-covid\\ changes\end{tabular}} & \begin{tabular}[c]{@{}l@{}}Yes, I use a safer messaging \\ app now\end{tabular} & 27 & 13 & 40 \\
 & \begin{tabular}[c]{@{}l@{}}Yes, I didn’t use any \\ messaging app before, but \\ do use it now\end{tabular} & 31 & 13 & 44 \\
 & \begin{tabular}[c]{@{}l@{}}No, I did use it before and \\ use it now\end{tabular} & 105 & 60 & 165 \\
 & \begin{tabular}[c]{@{}l@{}}No, I didn’t use it before and \\ don’t use it now\end{tabular} & 22 & 2 & 24 \\
Total &  & 193 & 89 & 282 \\ \hline
\end{tabular}
\vspace{0.8em}
\caption{Sector * post-covid changes crosstabulation}
\label{tab:sector*covid}
\end{table}

@@

Answer to these questions are plotted in Figure \ref{figure:chatbots_types} grouped by gender, sector and experience. As it was a multiple choice question, the counts are over the total number of people that answered. One of the most interesting questions that can arise from the plots, and the should require further study, is: Are teachers really aware of the possibilities of chatbots in the classroom? The reason to ask ourselves this question is due to the fact of the small number of responses for other uses in comparison with the other approaches proposed. Another way to look about it is that, unless you provide the specific instructions on how to use a tool, the teaching staff is already too busy or burn out to think about innovation.

@@

From a qualitative point of view, these results show consensus in a few key aspects:
% Maybe move to conclusions 
\begin{itemize}
 \item Teachers want to have support from their corresponding institution. This is probably due to the strict European data regulations and to avoid the overhead required to sing up and deleting students from year to year.

@@

@@

@@

% Emotial, mood for future work... Chatbots perform sentiment analysis

>>>>

For answering RQ1, teachers were queried about whether they use messaging apps in their classes, specifically, Telegram, WhatsApp, Slack, or any other messaging app, and whether they use messaging apps provided by their academic institution (see Table \ref{tab:use_messapps}). Overall, the majority of the teachers responded that they do use messaging apps in their classrooms, from which apps provided by the academic institution (n=159, 56\%) and WhatsApp (n=124, 44.0\%) were the most common responses. Only 19 teachers (6.7\%) replied that they do not use any messaging app in their class.

\begin{table}[h!tbp]
\begin{tabular}{lllll}
\hline
\multicolumn{1}{c}{Messaging App} & \multicolumn{2}{c}{Yes} & \multicolumn{2}{l}{No} \\ \cline{2-5} 
\multicolumn{1}{c}{} & \multicolumn{1}{c}{Frequency} & \multicolumn{1}{c}{\%} & Frequency & \multicolumn{1}{c}{\%} \\ \hline
Telegram & 62 & 22.0 & 220 & 78.0 \\
WhatsApp & 124 & 44.0 & 158 & 56.0 \\
Slack & 15 & 5.3 & 267 & 94.7 \\
Other & 60 & 21.3 & 222 & 78.7 \\
\begin{tabular}[c]{@{}l@{}}Provided by the Academic\\ Institution\end{tabular} & 159 & 56.4 & 123 & 43.6 \\
None & 19 & 6.7 & 263 & 93.3 \\ \hline
\end{tabular}
\vspace{0.8em}
\caption{Use of messaging apps in class}
\label{tab:use_messapps}
\end{table}

A chi-square test of independence was performed to examine the relation between instructors’ discipline, sector, and gender and any potential changes that occurred to the use of messaging apps due to the covid-19 pandemic. The relation between the latter variable and the instructors’ sector was significant, $X^{2}_{(4, N = 282)} = 9.598$, $p =0.048$. The frequencies cross tabulated are given in Table \ref{tab:sector*covid}. In effect, this finding indicates that how teachers responded to the use of messaging apps during the pandemic and in particular, whether they changed their habits in the use of apps for teaching purposes, was related to their sector (university vs vocational).
\begin{table}[h!tbp]
\begin{tabular}{llccc}
\hline
 & \multicolumn{1}{c}{} & \multicolumn{2}{c}{Sector} & Total \\ \cline{3-5} 
 & \multicolumn{1}{c}{} & \begin{tabular}[c]{@{}c@{}}University \\ (f)\end{tabular} & \begin{tabular}[c]{@{}c@{}}Vocational\\ (f)\end{tabular} &  \\ \cline{2-5} 
\multirow{4}{*}{\begin{tabular}[c]{@{}l@{}}Post-covid\\ changes\end{tabular}} & \begin{tabular}[c]{@{}l@{}}Yes, I use a safer messaging \\ app now\end{tabular} & 27 & 13 & 40 \\
 & \begin{tabular}[c]{@{}l@{}}Yes, I didn’t use any \\ messaging app before, but \\ do use it now\end{tabular} & 31 & 13 & 44 \\
 & \begin{tabular}[c]{@{}l@{}}No, I did use it before and \\ use it now\end{tabular} & 105 & 60 & 165 \\
 & \begin{tabular}[c]{@{}l@{}}No, I didn’t use it before and \\ don’t use it now\end{tabular} & 22 & 2 & 24 \\
Total &  & 193 & 89 & 282 \\ \hline
\end{tabular}
\vspace{0.8em}
\caption{Sector * post-covid changes crosstabulation}
\label{tab:sector*covid}
\end{table}
The majority of the teachers (n=165) mentioned that no changes in their habits occured due to the emergency remote teaching, as the use of messaging apps was part of their teaching practices and remains the same, from which 105 teachers come from the university, and 60 teachers from the vocational sector. The relationships between changes in the use of messaging apps, due to the covid-19 pandemic, and the instructors’ discipline, $X^{2}_{(24, N = 282)} = 44.856$, $p =0.006$, as well as the gender, $X^{2}_{(8, N = 282)} = 16.249$, p= 0.039 were also significant. This finding indicates that how teachers responded to the use of messaging apps during the pandemic was also related to their gender and discipline. In fact, from the majority of the teachers who did not change their habits in this respect (n=165), most of them are males (n=114), and come from the technology (n=60) and engineering (n=53) disciplines (see Tables \ref{tab:gender*covid}, \ref{tab:discipline*covid}).

\begin{table}[h!tbp]
\begin{tabular}{llcccc}
\hline
 & \multicolumn{1}{c}{} & \multicolumn{3}{c}{Gender} & Total \\ \cline{3-6} 
 & \multicolumn{1}{c}{} & \begin{tabular}[c]{@{}c@{}}Male \\ (f)\end{tabular} & \begin{tabular}[c]{@{}c@{}}Female\\ (f)\end{tabular} &  \begin{tabular}[c]{@{}c@{}}PNTS\\ (f)\end{tabular} & \\ \cline{2-6} 
\multirow{4}{*}{\begin{tabular}[c]{@{}l@{}}Post-covid\\ changes\end{tabular}} & \begin{tabular}[c]{@{}l@{}}Yes, I use a safer messaging \\ app now\end{tabular} & 20 & 20 & 0 & 40 \\
 & \begin{tabular}[c]{@{}l@{}}Yes, I didn’t use any \\ messaging app before, but \\ do use it now\end{tabular} & 20 & 23 & 1 & 44 \\
 & \begin{tabular}[c]{@{}l@{}}No, I did use it before and \\ use it now\end{tabular} & 114 & 47 & 4 & 165 \\
 & \begin{tabular}[c]{@{}l@{}}No, I didn’t use it before and \\ don’t use it now\end{tabular} & 18 & 6 & 0 & 24 \\
 & Other & 7 & 2 & 0 & 9 \\
Total &  & 179 & 98 & 5 & 282 \\ \hline
\end{tabular}
\vspace{0.8em}
\caption{Gender * post-covid changes crosstabulation}
\label{tab:gender*covid}
\end{table}

\begin{table}[h!tbp]
\centering
\resizebox{\textwidth}{!}{
\begin{tabular}{llcccccccc}
\hline
 & \multicolumn{1}{c}{} & \multicolumn{7}{c}{Disciplines} & Total \\ \cline{3-10} 
 & \multicolumn{1}{c}{} & \begin{tabular}[c]{@{}c@{}}Engi- \\ neering(f)\end{tabular} & \begin{tabular}[c]{@{}c@{}}Social \\ Sciences (f)\end{tabular} &  \begin{tabular}[c]{@{}c@{}}Sciences\\ (f)\end{tabular} & \begin{tabular}[c]{@{}c@{}}Bio-\\ medicine (f)\end{tabular} & \begin{tabular}[c]{@{}c@{}}Human-\\ ities(f)\end{tabular} & \begin{tabular}[c]{@{}c@{}}Technol-\\ ogy(f)\end{tabular} &  \begin{tabular}[c]{@{}c@{}}Other\\ (f)\end{tabular} & \\ \cline{2-10} 
\multirow{4}{*}{\begin{tabular}[c]{@{}l@{}}Post-covid\\ changes\end{tabular}} & \begin{tabular}[c]{@{}l@{}}Yes, I use a safer \\ messaging app now\end{tabular} & 7 & 11 & 5 & 5 & 1 & 10 & 1 & 40 \\
 & \begin{tabular}[c]{@{}l@{}}Yes, I didn’t use any \\ messaging app before, \\ but do use it now\end{tabular} & 13 & 12 & 5 &  4 &  2 &  7 & 1 & 44 \\
 & \begin{tabular}[c]{@{}l@{}}No, I did use it before \\ and use it now\end{tabular} &60 &  26 &  20 &  4 & 0 & 53 &  2 &  165\\
 & \begin{tabular}[c]{@{}l@{}}No, I didn’t use it \\ before and don’t \\ use it now\end{tabular} & 14 &  4 & 1 & 2 & 1 & 1 & 1 & 24 \\
 & Other & 5  & 0 & 1 & 1 & 1 & 1 & 0 & 9 \\
Total & & 99 &  53&  32&  16&  5&  72&  5&  282\\ \hline
\end{tabular}}
\vspace{0.8em}
\caption{Discipline * post-covid changes crosstabulation}
\label{tab:discipline*covid}
\end{table}

Although gender differences are not significant, female teachers answered they use instant messaging apps more than male teachers (about 10\% more). Also, teachers in vocational Education also use Whatsapp more than university teachers.
With respect to the distribution of the use of messaging apps per age, there are no significant differences for Whatsapp and apps provided by their institutions. % Olia requests support for this.
However, younger teachers also use Telegram with more than 25\%, a percentage that falls to about 10\% for teachers that are 55 or older. The third column compares the use of these apps considering the number of years of experience. One interesting result is that, contrary to the idea of technology adaptation issues when considering older teachers, about 65\% of teachers with more than 25 years of experience use the platforms provided by their institutions while the percentage goes down to less than 50\% for teachers with 6-15 years of experience.

@@

Some of the questions in the survey were focused on the impact of the COVID19 pandemic between 2020 and 2021, in the teachers' viewpoints towards the use of instant messaging apps int their class. The 282 
% Fran, please check Olia's comment as having 282 answers -> I did remove some responses that Olia considered 'No' by default. I' ll keep it to 282 to be consistent
answers are summarized in Fig. \ref{figure:chatbots_use_postcovid}, showing that about 77\% of teachers already used these tools before the pandemic and kept using them during the pandemic lockdowns that forced students and educators to use remote education schemes. Moreover, approximately 15\% of them switched their messaging app for one that offered a safer interaction with their students. According to the responses, an additional 16\% started using messaging apps during the pandemic for the first time in their classes. Aggregating the answers, currently 91\% of the teachers are using messaging apps as learning tools.

@@

\begin{figure}[h!tbp]
\begin{center}
    \begin{minipage}[t]{0.85\textwidth}
        \centering
<<gender_covid, echo=FALSE, fig.width=19/2.54, fig.height=5.5/2.54>>=
survey.pilot.postCOVID <- within(survey.pilot.postCOVID, PostCOVID.changes <- factor(PostCOVID.changes, levels=names(sort(table(PostCOVID.changes), decreasing=TRUE))))
ggplot(survey.pilot.postCOVID, aes( x = PostCOVID.changes, y = ..count.., group=Gender, fill = Gender ) ) + geom_bar()+ theme(axis.text.x = element_text(angle = 90)) +coord_flip()+ theme_tufte()+ theme(axis.title.y=element_blank())
@
%       \includegraphics[width=\textwidth]{figures/gender_covid.pdf}
    \end{minipage}
    \begin{minipage}[t]{0.85\textwidth}
        \centering
<<sector_covid, echo=FALSE, fig.width=19/2.54, fig.height=5.5/2.54>>=
ggplot(survey.pilot.postCOVID, aes( x = PostCOVID.changes, y = ..count.., group=Sector, fill = Sector ) ) + geom_bar()+ theme(axis.text.x = element_text(angle = 90)) +coord_flip()+ theme_tufte()+ theme(axis.title.y=element_blank())
@
%       \includegraphics[width=\textwidth]{figures/sector_covid.pdf}
    \end{minipage}
    \begin{minipage}[t]{0.85\textwidth}
        \centering
<<experience_covid, echo=FALSE, fig.width=19/2.54, fig.height=5.5/2.54>>=
ggplot(survey.pilot.postCOVID, aes( x = PostCOVID.changes, y = ..count.., group=Experience, fill = Experience ) ) + geom_bar()+ theme(axis.text.x = element_text(angle = 90)) +coord_flip()+ theme_tufte()+ theme(axis.title.y=element_blank())
@       
%       \includegraphics[width=\textwidth]{figures/experience_covid.pdf}
    \end{minipage}
\end{center}
    \caption{Impact of COVID in the use of messaging apps in University and vocational education.}
    % Have different sub-captions so that we can refer to specific subfigures - Olia
\label{figure:chatbots_use_postcovid}
\end{figure}

%A chi-square test of independence was performed to examine the relation between instructors’ discipline, sector, and gender and any potential changes that occurred to the use of messaging apps due to the covid-19 pandemic. The relation between the latter variable and the instructors’ sector was significant, $X^{2}_{(4, N = 282)} = 9.598$, $p =0.048$. The frequencies cross tabulated are given in Table \ref{tab:sector*covid}. In effect, this finding indicates that how teachers responded to the use of messaging apps during the pandemic and in particular, whether they changed their habits in the use of apps for teaching purposes, was related to their sector (university vs vocational). 
The relationships between changes in the use of messaging apps, due to the covid-19 pandemic, and the instructors’ discipline, $X^{2}_{(24, N = 282)} = 44.856$, $p =0.006$, as well as the gender, $X^{2}_{(8, N = 282)} = 16.249$, $p= 0.039$ were also significant.

@@

Answer to these questions are plotted in Figure \ref{figure:chatbots_types} grouped by gender, sector and experience. As it was a multiple choice question, the counts are over the total number of people that answered.

@@

@@

% Add something about intro
% RQ1 discussion about messaging app: they consider a IM the Moodle messaging
% Olia: it's not so easy to change habits, even in the case of major crisis. - JJ
% Relationship between changes, gender and discipline (Olia) - JJ

% Olia: we have something for RQ 2 and 4.
 One of the most interesting questions that can arise from the plots, and the should require further study, is: Are teachers really aware of the possibilities of chatbots in the classroom? The reason to ask ourselves this question is due to the fact of the small number of responses for other uses in comparison with the other approaches proposed. Another way to look about it is that, unless you provide the specific instructions on how to use a tool, the teaching staff is already too busy or burn out to think about innovation.

% Noorhan comments negative experience with customer service chatbots, in which case people will not see the value of using it in their classes. They are not aware of something like that. Teaching chatbots are difficult to develop. Using very simple chatbots, or simply conversational interfaces, could help introduce this technology - JJ

From a qualitative point of view, these results show consensus in a few key aspects:
\begin{itemize}
 \item Teachers want to have support from their corresponding institution. This is probably due to the strict European data regulations and to avoid the overhead required to sing up and deleting students from year to year.

% Rq4: they seem to prefer interaction with students - Olia 
% Alberto: using new metric so grade students, and also abotu horizontality

@@

% Noorhan suggest that using some simple things could increase the perceived usefulness - JJ

@@

% Emotinal, mood for future work... Chatbots perform sentiment analysis
